---
title: "Food Production in Kenya"
author: "Florence Muriuki"
date: "2019-07-21"
output: html_document
---


*The purpose of this project is to showcase use of R shiny in creating simple interactive dashboards. I decided 
to use food production dataset because in future I would like to dig deeper into why we experience food shortages 
in some parts of Kenya.*


We all have our favourite foods. Some people can't wait for their favourite fruit season and 
are willing to pay exorbitant amounts during the off season. In the recent past, there has been 
a craze around avocadoes to an extent that avocadoes sold at above Ksh 40  a piece in some parts of Nairobi. Using data from [AFA](https://www.agricultureauthority.go.ke/?page_id=976), I looked at 
avocado production within the country between 2012 and 2017. Please note that data from 2015 is missing.


```{r, include = F}
library(ggplot2)
library(ggmap)
library(plotly)
```

```{r, include = F}
food = readRDS("Food/Cleaned-Data/food_production_2012_2017.rds")
county_shapefiles = readRDS("Food/Cleaned-Data/county_shapefiles.rds")

```

```{r, include = F}
avocadoes = food %>%
  dplyr::filter(crop_type == "Avocadoes") %>%
  group_by(year) %>%
  summarise(produce = sum(volume_in_mt), area = sum(area_in_ha))

```

### Trend in Avocado Production 
```{r, echo = F, fig.width = 12, fig.height = 6}
g <- ggplot(avocadoes, aes(x = year, y = produce)) + geom_line(lwd = 2, col = "#006400") +            
        ggtitle(label = paste("Avocadoes Annual Produce in Tonnes")) +
        theme(axis.ticks = element_blank(), 
              axis.title = element_blank(),
              axis.text = element_text(size = 14,colour = "#006400"), 
              plot.title = element_text(size = 16, colour = "#006400", face = "bold")) + 
        scale_x_continuous(breaks = seq(min(avocadoes$year, na.rm = T), max(avocadoes$year, na.rm = T), 1),
                           labels = seq(min(avocadoes$year, na.rm = T), max(avocadoes$year, na.rm = T), 1)) +
        scale_y_continuous(labels = scales::comma)
ggplotly(g)

```

Avocadoes production has been rising since 2012 and the growth is likely to continue due to increased local and export market.


```{r, echo = F}
counties = food %>%
  dplyr::filter(crop_type == "Avocadoes", year == 2017)
```

```{r, echo = F, fig.width = 6, fig.height = 6 }

county_shapefiles1 <- county_shapefiles %>%
        dplyr::left_join(counties, by = "county") %>%
        dplyr::rename("produce" = volume_in_mt)

g <- ggplot(county_shapefiles1, aes(x = long, y = lat, group = group, label = county)) +
  geom_polygon(aes(fill = produce)) +
  theme_classic() + theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(), 
    axis.title = element_blank(),
    axis.line = element_blank(),
    plot.title = element_text(size = 16, colour = "#006400", face = "bold")) +
   guides(fill = guide_legend(title = "Produce")) +
  ggtitle(label = paste("Avocadoes Production in Tonnes at County Level")) +
  scale_fill_gradient(low = "#7CFC00", high = "#006400")
ggplotly(g) 
 

```
*The darker the shade, the higher the produce.*

 
 As of 2017, Murang'a was the leading producer of avocadoes.
 
 To see trends for other crops and what each county is producing, visit [food production dashboard](https://florence.shinyapps.io/food_production/.)
 
 
 
 Data and code is [here](https://github.com/florencewambui/food_in_kenya)
 
 Data source: [AFA](https://www.agricultureauthority.go.ke/?page_id=976)
 
 Other references: 

 * [Standard Media](https://www.standardmedia.co.ke/article/2001326301/china-s-avocado-export-boom)
 
 * [Business Daily](https://www.businessdailyafrica.com/news/Avocado-exports-banned-as-price-rises/539546-4272600-9gthm2z/index.html)

Tools: 

* [Tabula](https://tabula.technology/) for data tables extraction from pdfs

* [R](https://www.r-project.org/)

       
       

